Spectral density estimation using P-spline priors

### 2019-04-18 pSplines.tex
format change

\documentclass[aps,reprint,amsmath,amssymb,showpacs,showkeys]{revtex4-1}% APS journal style
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{hyperref}% add hypertext capabilities
\usepackage{natbib}  % Package for Bibtex
%\usepackage{amsmath}  % Package for typesetting formulae
%\usepackage{amssymb}  % Package for some math symbols and fonts
\usepackage[justification=justified]{caption}
%\usepackage[caption=false]{subfig}
\usepackage{subcaption} % subfigure
\usepackage{amsopn}
\usepackage{float}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}
\tikzstyle{sqr2} = [rectangle, minimum width=3.5cm, minimum height=0.9cm, text centered, draw=black, fill=yellow!30]

\usepackage{ulem} % \sout function
\usepackage{booktabs} % Top and bottom rules for tables

%\usepackage{breqn}

\newcommand{\bt}[1]{\ensuremath{\mathbf{#1}}}  % bold math
%\newcommand{\bm}[1]{\mbox{\boldmath $#1$}}
\newcommand{\cb}{ \color{blue}}
\newcommand{\cred}{\color{red}}
\newcommand{\jv}[1]{\textcolor{cyan}{JV: #1}}
\newcommand{\Msun}{$\text{M}_{\odot}$}

\begin{document}
\graphicspath{{figures/}}

% Define author, title, and date
\title[Spectral density estimation using P-spline priors]{Spectral density estimation using P-spline priors}

\author{Patricio Maturana-Russel$^{1}$ and Renate Meyer$^1$}
\affiliation{$^1$ Department of
  Statistics, University of Auckland, Auckland 1142, New Zealand}

\begin{abstract}
This article proposes a P-spline prior to estimating the spectral density of a stationary time series.  Our proposal is a particular case of the B-spline prior algorithm.  We assess our proposal via a simulation study and two real data sets.
\end{abstract}

\pacs{04.30.-w, 02.50.-r, 05.45.Tp, 97.60.Bw}

\keywords{P-spline prior, B-spline prior, spectral density estimation, Whittle likelihood}

@@

\section{Introduction}
The power spectral density (PSD), or simply spectral density, describes the distribution of the power or variance for each individual frequency composing a time series.  It embodies useful information for the study of stationary time series.  An absolutely summable autocovariance function, that is $\sum_{h=-\infty}^{\infty} |\gamma(h)| <\infty$, guaranties the existence of the spectral density function of a zero-mean weakly stationary time series, being continuous, bounded and given by
\begin{align*}
f(\lambda) = \dfrac{1}{2\pi} \sum_{h = -\infty}^{\infty}\gamma(h)\exp\left(-i h \lambda \right), 
\end{align*}
where $-\pi < \lambda \leq \pi$ is the angular frequency.

Methods to estimate this function go from parametric to nonparametric approaches.  The former are basically based on autoregressive moving average (ARMA) models.  As any parametric model, the results tend to be biased if the model approximation to the underlying time series is poor.

Most of the nonparametric methods work by smoothing the periodogram
\begin{align*}
I_n(\lambda) = \dfrac{1}{2 \pi n} \left| \sum_{t=1}^{n} Y_t \exp \left( -i t \lambda\right)\right|^2,
\end{align*}
where $-\pi < \lambda \leq \pi$ is the angular frequency and $Y_t$ is a stationary time series of length $n$, because this fluctuates around the true spectral density function.  This function has interesting features, for instance, it is asymptotically unbiased and follows an exponential distribution with mean $f(\lambda)$.  However, it is not consistent \citep{Brockwell:1986}.

Bayesian nonparametric approaches for spectral density estimation allow the data to determine the complexity of the model.  These approaches avoid the comparison of models which vary in complexity and adapt a single model according to the data. \cite{Gangopadhyay:1999} fitted a fixed low order piecewise polynomials to the log-spectral density function, allowing free known number and location of knots.  For this, a reversible jump Markov chain Monte Carlo (RJMCMC) \citep[RJMCMC;][]{Green:1995} algorithm was used.

\cite{Choudhuri:2004} allocated a Bernstein polynomial as prior on the spectral density function.  This prior is a mixture of beta densities weighted by a Dirichlet process.  The number of components, which is a smoothing parameter, has a discrete prior.  \cite{Edwards2018} extended the method replacing the beta densities for B-spline functions.  These functions have the advantage of having local support, unlike the beta densities, which allow more flexibility and have the potential of modelling spectral densities with sharp and abrupt changes.  Thus, the B-spline prior outperforms the Bernstein polynomial in estimating complex PSDs \citep{Edwards2018}.  However, this flexibility comes with a much higher computational cost.

The Bernstein polynomial and the B-spline methods make use of Whittle's approximation to the Gaussian likelihood, known simply as Whittle likelihood \citep{Whittle:1957}.  One of the advantages of this approximation is that involves directly the spectral density, unlike the true likelihood function.

Taking advantage of the limiting independent exponential distribution of the periodogram, the Whittle likelihood for a mean centred weakly stationary time series $\bm{Y}$ of length $n$ is defined as
\begin{align}
 L(\bm{Y} | f) &= \prod_{l=1}^{\nu}\dfrac{1}{f(\lambda_{l})} e^{-I_n(\lambda_l)/f(\lambda_l)}  \nonumber \\
 &\propto \exp \left( - \sum_{l=1}^{\nu} \log f(\lambda_l) + \dfrac{I_{n}(\lambda_l)}{f(\lambda_l)} \right), \label{eq:whittle_like}
\end{align}
where $\lambda_l = 2\pi l / n$ are the Fourier frequencies, $f(\cdot)$ spectral density of $\bm{Y}$, and $\nu = \left\lfloor (n-1)/2 \right\rfloor$ is the greatest integer value less than or equal to $(n-1)/2$.

Motivated by the flexibility of the B-spline prior and its associated high computational cost, in this work, we implement and test a particular case of the B-spline prior.  For this, we fix the number of B-spline densities and place a P-spline prior on the spectral density instead.  This approach allows to penalize automatically the inclusion of B-spline densities and save significantly computational time without sacrificing accuracy.  Actually, as will show, it outperforms the B-spline prior in our tests.  In addition, we propose a knot location scheme based on the periodogram, which aims to deal with PSD shapes that possess  abrupt peaks, without increasing computational time.  We show in an example how this scheme improves significantly the estimates.

We start defining the B-spline densities and the B-spline prior. Then, we define the P-spline prior and discuss its implementation.  Finally, we test our approach and compare it to the B-spline prior in a simulation study, and two real problems.

\section{B-spline prior}

A mixture of beta densities with only integer parameters can approximate uniformly any continuous density on $[0,1]$.  Let $G$ be a cumulative distribution function (cdf), with continuous density $g$, then 
\begin{align*}
\widehat{G}(\omega) &= \sum_{k=1}^{K} G \left( \dfrac{k-1}{K} , \dfrac{k}{K} \right] \beta(\omega; k, K-k+1)\\
&= \sum_{k=1}^{K} w_k \beta(\omega; k, K-k+1)
\end{align*}	  
converges uniformly to $G(\omega)$, where $G(u,v] = G(v) - G(u)$ and $\beta(\omega; a,b)$ is the beta density with parameters $a$ and $b$.
The Bernstein polynomial prior, which is used to describe a nonparametric prior for probability densities in the unit interval, is based on this approximation \citep{Petrone:1999a,Petrone:199b}.  \cite{Choudhuri:2004} proposed it as a prior on the spectral density function.

As shown by \cite{Perron:2001}, the mixture of beta distributions is unable to provide a full coverage of the space of cdfs on $[0,1]$.  They defined a loss function and showed that this cannot be arbitrarily small under this approach.  Alternatively, they showed that if the beta distributions are replaced by B-spline distributions of fixed order (shown for order~2) with variable knots, the loss function can be arbitrarily small by increasing the number of knots.  The adequacy of the approximation is a function of the number of knots.  This work motivated \cite{Edwards2018} to propose the B-spline prior on the spectral density function.

@@

The \textit{B-spline densities}, normalized B-spline functions, are defined by
\begin{align*}
	b_{k,r}(\omega; \bm{\xi}) = \left(r+1\right) \dfrac{B_{k,r}(\omega;\bm{\xi})}{\xi_{k+r} - \xi_{k-1}}.
\end{align*}

@@

\subsection*{Prior description}
The B-spline prior is a mixture of B-spline densities defined as
\begin{align}
s_r(\omega; K,\bm{w}_K, \bm{\xi}) = \sum_{k=1}^{K} w_{k,K} b_{k,r}(\omega;\bm{\xi}),
\label{eq:bs_prior}	
\end{align}
where $K$ is the number of B-spline densities  $b_{k,r}(\cdot)$ of fixed degree $r$, $\bm{w}_K=(w_{1,K},\cdots, w_{K,K})$ the weight vector, and $\bm{\xi}$ is the knot sequence.

@@

Assuming that the weights are induced by a cdf $G$ on $[0,1]$, \cite{Choudhuri:2004} proposed a Dirichlet process as a prior for the latter.  This approach avoids the potential difficulties caused by the parameter dimension changes associated to RJMCMC.  \cite{Edwards2018} followed this approach, but in addition they assumed that the $K-r$ internal knot differences are induced by a cdf $H$ on $[0,1]$, for which was allocated another Dirichlet prior, that is
\begin{align*}
\Delta_{j} = \xi_{j+r} - \xi_{j+r-1} = H\left(\dfrac{j-1}{K-r}, \dfrac{j}{K-r} \right],
\end{align*}
for $j=1,\dots,K-r$.

Therefore, the B-spline prior on the spectral density function parametrized in terms of $K$, $G$, and $H$ is given by
\begin{align*}
s_r(\omega;K,G,H) = \sum_{k=1}^{K} G \left( \dfrac{k-1}{K} , \dfrac{k}{K} \right] b_{k,r}(\omega;H).
\end{align*}

The B-spline prior is a mixture of B-spline densities with local support, unlike the Bernstein polynomial prior which uses beta densities with full support on the unit interval.  This feature allows the B-spline method more flexibility for spectral density estimation \citep{Edwards2018}.

When there are no internal knots, the B-spline basis turn into the Bernstein polynomial basis.  Therefore, the B-spline prior can be considered as a generalization of the Bernstein polynomial prior.

@@

The prior discussed above has been defined for the unit interval $[0,1]$.  However, the spectral density function $f(\cdot)$ is defined on the interval $[0,\pi]$, therefore its prior is reparametrized and defined as
\begin{align*}
f(\pi \omega) = \tau \times s_r(\omega;K,G,H), \quad \omega \in [0,1],
\end{align*}
where $\tau = \int_{0}^{1}f(\tau \omega)\text{d}\omega$ is the normalizing constant.

To sum up, the prior has the following components:
\begin{itemize}
	\item $G \sim \text{DP}(M_G, G_0)$, where $M_G>0$ is the concentration parameter and $G_0$ is the base distribution.  It produces the weights for the $K$ B-spline densities.  
	\item $H \sim \text{DP}(M_H, H_0)$, where $M_H>0$ is the concentration parameter and $H_0$ is the base distribution.  It produces the knot locations determining the shape and location of the B-spline densities.
	\item $p(K) \propto \text{exp}(-\theta_K K^2)$ for $K=1,\dots,K_{\text{max}}$.  $K$ is the number of B-spline densities used in Eq. \eqref{eq:bs_prior}.  It plays the role of smoothing parameter, that is smaller values implies smoother spectral density estimates.  $K_{\text{max}}$ must be limited  for computational reasons.  In practice, pilot runs are used to determine it.
	\item $\tau \sim \text{Inverse-Gamma}(\alpha_{\tau},\beta_{\tau})$ is the normalizing constant.  $\alpha_{\tau}$ and $\beta_{\tau}$ are the shape and scale parameters, respectively.
\end{itemize}

\subsection*{Implementation discussion}

The B-spline prior proposed by \cite{Edwards2018} involves 2 Dirichlet processes placed on $G$ and $H$, the weights of the B-spline densities and the knot locations, respectively.  The sampling from them is carried out by the stick-breaking construction \citep{Sethuraman:1994}.  This method consists basically in a representation of the Dirichlet process as a sequential division of the unit interval infinite times according to a beta distribution.  For computational reasons, the number of components of this representation of $G$ and $H$ is limited to a large but finite positive integer, that is $L_G$ and $L_H$, respectively.  These parameters control the quality of the approximations.  Larger values improve the approximation, but increase the number of calculations and consequently the computation time.

The stick-breaking approximation requires the re-parametrization of $G$ to $(Z_0,Z_1,\dots, Z_{L_G},V_1,\dots,V_{L_G})$, which are all independent, such that
\begin{align*}
G&=\sum_{l=1}^{L_G} p_l \delta_{Z_l} + \left(1- \sum_{l=1}^{L_G} p_l\right)\delta_{ Z_{0}},
%p1 &= V_1, \\
%p_l&=\left( \prod_{j=1}^{l-1} (1 - V_j) \right) V_l, \quad l \geq 2\\
%p_0&=1-\sum_{l=1}^{L_G}p_l,\\
%V_l &\sim \text{Beta}(1,M_G), \quad l=1,\dots,L_G,\\
%Z_l &\sim G_0, \quad l = 0,1\dots,L_G.
\end{align*}
where $p_1 = V_1$, 
\begin{align*}
p_l &= V_l \prod_{j=1}^{l-1} (1 - V_j), \:\: \text{for } l \geq 2, \\
p_0 &= 1-\sum_{l=1}^{L_G}p_l,
\end{align*}
$\delta_{Z_{0}}$ is a degenerate probability density, that is $\delta_{Z_0}=1$ at $Z_0$ and 0 otherwise, $V_l \sim \text{Beta}(1,M_G)$, for $l=1,\dots,L_G,$ and $Z_l \sim G_0$, for $l = 0,1\dots,L_G$.  Conditional on $K$, the weights of the B-spline densities are defined as
\begin{align*}
	w_{k,K} = \sum_{l=0}^{L_G}p_l I\left\{ \dfrac{k-1}{K} < Z_l < \dfrac{k}{K}\right\}.
\end{align*}

Analogously, $H$ is re-parametrized as $(X_0,\dots,X_{L_H}, U_1,\dots,U_{L_H})$ by the stick-breaking method truncated at $L_H$.  Thus, the knot differences and consequently the knot locations are defined as
\begin{align*}
\Delta_{j} &= \xi_{j+r} - \xi_{j+r-1} \\
&= \sum_{l=0}^{L_H}q_l I\left\{ \dfrac{j-1}{K-r} < X_l < \dfrac{j}{K-r} \right\},		
\end{align*}
where $q_l$ is defined similarly to $p_l$ that was described above.

As a result, the joint prior for the parameter vector $\bm{\theta} = (\textbf{v},\textbf{z},\textbf{u},\textbf{x},K,\tau)$ is given by
\begin{align}
p(\bm{\theta}) \propto & \left( \prod_{l=1}^{L_G} M_G(1-v_l)^{M_G -1} \right) \left( \prod_{l=0}^{L_G}g_0(z_l) \right) \nonumber \\
&\times \left( \prod_{l=1}^{L_H} M_H(1-u_l)^{M_H -1} \right) \left( \prod_{l=0}^{L_H}h_0(x_l) \right) \nonumber \\
&\times p(K) p(\tau) \label{eq:prior}
\end{align}
An unnormalized joint pseudo-posterior is obtained by multiplying this prior density and the Whittle likelihood defined in Eq. \eqref{eq:whittle_like}.

The sampling is performed via the Gibbs sampling technique.  More details about the MCMC implementation of the B-spline prior method can be found in \cite{Edwards2018}.\\

As shown by \cite{Edwards2018}, the B-spline prior outperforms the Bernstein polynomial prior in estimating spectral densities with sharp and abrupt changes. In the case of simple ones, there is no a significant difference.  However, in practice, this function is unknown, which makes necessary the use of general methods, such as the B-spline prior algorithm.

The Bernstein polynomial prior requires the calculation of the beta densities only once and they can be re-utilized across MCMC steps.  On the other hand, the B-spline prior method requires the calculation of the B-spline densities at each iteration, since these vary according to the knot locations and the number of B-spline densities.  Moreover, the Dirichlet processes approximated via the stick-breaking method require $2\!\times\!(L_G + L_H)$ calculations (see Eq. \eqref{eq:prior}) at each iteration.  All of this makes the B-spline prior algorithm much more complex in comparison to the Bernstein polynomial prior.

The complexity of the B-spline prior algorithm is mainly because it allows variable number of B-spline densities $K$.  This has prompted us to the implementation and assessment of the P-spline prior algorithm, which uses a fixed number of B-spline densities and knot locations, avoiding the Dirichlet processes and the recalculation of the B-spline densities.  One of its advantages is that this method keeps the local support of the B-spline densities, unlike the Bernstein polynomials, which can potentially allow to estimate spectral densities with sharp and abrupt changes.

@@

Fixing the number of B-spline densities $K$ in Eq. \eqref{eq:bs_prior}, the prior for the unit interval $[0,1]$ is defined as
\begin{align*}
	s_r(\omega;\bm{w}, \bm{\xi}) = \sum_{k=1}^{K} w_{k} b_{k,r}(\omega;\bm{\xi}),	
\end{align*}
where $\bm{w} = (w_1,\dots,w_K)$, and $\bm{\xi}$ contains equidistant internal knots on $[0,1]$.  Therefore, the P-spline prior on the spectral density function is given by
\begin{align*}
f(\pi \omega) = \tau \times s_r(\omega;\bm{w}, \bm{\xi}), \quad \omega \in [0,1],
\end{align*}
where $\tau = \int_{0}^{1}f(\tau \omega)\text{d}\omega$ is the normalizing constant.

The roughness penalty in a frequentist context is carried out by penalizing the likelihood function.  In a Bayesian context, this is translated into the prior distribution \citep{Lang:2004} for the $d^{\text{th}}$ order difference of successive weights, known also as B-spline parameters.

>>>>

%\documentclass[11pt,a4paper]{article} % Document specs
\documentclass[twocolumn,final]{svjour3}

% Useful packages
\usepackage{graphicx}  % Package for importing postscript images
\usepackage[authoryear, round, numbers]{natbib}  % Package for Bibtex
\usepackage{amsmath}  % Package for typesetting formulae
\usepackage{hyperref}  % Package for hyperlinking
%\usepackage{Sweave}  % Package for Sweave
\usepackage[top=1.5in, bottom=1.5in, left=1in, right=1in]{geometry}
%\usepackage{caption}
\usepackage{amsfonts}

\usepackage{booktabs} % Top and bottom rules for tables
%\usepackage{lmodern}

\newcommand{\bm}[1]{\mbox{\boldmath $#1$}}
\newcommand{\w}{{\bf w}}
% Define author, title, and date

\title{Spectral density estimation using P-spline priors}
\author{Patricio Maturana-Russel \and Renate Meyer}

\institute{Patricio Maturana-Russel \at Department of Statistics, University of Auckland, Auckland, New Zealand \\ \email{p.russel@auckland.ac.nz} 
\and Renate Meyer  \at Department of Statistics, University of Auckland, Auckland, New Zealand 
}

%\date{} % Today's date appears if unspecified

\graphicspath{{figures/}}

\begin{document}\sloppy

@@

\begin{abstract}

This article proposes a Bayesian approach to estimating the spectral density of a stationary time series using a  P-spline prior.  Our proposal is motivated by the B-spline Dirichlet process prior of \cite{Edwards2018} combined with Whittle's likelihood with the aim of reducing the computational complexity of posterior computations. The strength of the B-spline Dirichlet process prior
over the Bernstein-Dirichlet process prior of \cite{Choudhuri:2004}  lies in its ability to estimate spectral densities with sharp peaks and abrupt changes due to the flexibility of B-splines with variable number and location of knots. Here, we fix the number and location of the knots and specify a P-spline prior.  
We demonstrate in a simulation study and two real  case studies that this approach retains the flexiblity of the B-splines, achieves the same ability
to accurately estimate peaks due to a new data-driven knot allocation scheme but significantly reduces the computational costs.

\end{abstract}

\keywords{P-spline prior \and B-spline prior \and spectral density estimation \and Whittle likelihood}

\section{Introduction}
The power spectral density (psd), or simply spectral density, describes the distribution of the power or variance over the individual frequency components of a time series.  It embodies useful information for the study of stationary time series.  For a zero-mean weakly stationary time series, an absolutely summable autocovariance function, i.e.\ $\sum_{h=-\infty}^{\infty} |\gamma(h)| <\infty$, guarantees the existence of a continuous and bounded spectral density function given by
\begin{align*}
f(\lambda) = \dfrac{1}{2\pi} \sum_{h = -\infty}^{\infty}\gamma(h)\exp\left(-i h \lambda \right), 
\end{align*}
where $-\pi < \lambda \leq \pi$ is the angular frequency.

Methods to estimate this function range from parametric to nonparametric approaches.  The former are mainly based on fitting  autoregressive moving average (ARMA) models. However, the corresponding spectral density will be biased if the ARMA model does not adequately describe the dependence structure of the time series. This is a downside of all parametric approaches.
They are efficient if the model fits well but because they make restrictive assumptions about the data-generating mechanism, their inference is sensitive to model misspecifications and may  be severely biased. 
Bayesian nonparametric approaches, on the other hand, put a prior distribution on the set of all possible densities that could have generated the data not just a small parametric subset thereof. They thus avoid the problem of choosing between different parametric models  by an intrinsic data-driven determination of the model complexity.

As the periodogram fluctuates around the true spectral density, most  nonparametric methods for spectral density estimation hinge on  smoothing the periodogram
\begin{align*}
I_n(\lambda) = \dfrac{1}{2 \pi n} \left| \sum_{t=1}^{n} Y_t \exp \left( -i t \lambda\right)\right|^2,
\end{align*}
in one way or another, where $-\pi < \lambda \leq \pi$ is the angular frequency and $\bm{Y}=(Y_1,\ldots,Y_n)$ is a stationary time series of length $n$. The periodogram has interesting features, for instance, it is asymptotically unbiased and follows an exponential distribution with mean $f(\lambda)$.  However, it is not consistent \citep{Brockwell:1986}.

Most Bayesian approaches to psd estimation use the so-called Whittle likelihood, introduced by \citep{Whittle:1957} as an approximation to the likelihood of a Gaussian stationary time series.  Based on the  limiting independent exponential distribution of the periodogram, the Whittle likelihood for a mean centered weakly stationary time series $\bm{Y}$ of length $n$ is defined as
\begin{align}
L(\bm{Y} | f) &= \prod_{l=1}^{\nu}\dfrac{1}{f(\lambda_{l})} e^{-I_n(\lambda_l)/f(\lambda_l)}  \nonumber \\
&\propto \exp \left( - \sum_{l=1}^{\nu} \log f(\lambda_l) + \dfrac{I_{n}(\lambda_l)}{f(\lambda_l)} \right)\label{eq:whittle_like}
\end{align}
where $\lambda_l = 2\pi l / n$ are the Fourier frequencies, $f(\cdot)$ spectral density of $\bm{Y}$, and $\nu = \left\lfloor (n-1)/2 \right\rfloor$ is the greatest integer value less than or equal to $(n-1)/2$. Advantages of this approximation is that it depends directly on the spectral density, unlike the true Gaussian likelihood function, and gives a good approximation for non-Gaussian time series under certain conditions as shown by \cite{ShaoXiaofeng2007ASTf}.

Bayesian nonparametric prior specifications for the psd include \cite{Cart:1997}
who used a smoothing prior on the log psd.  \cite{Rosen:2007} decompose the log psd into a linear and nonlinear component and  put a Gaussian process prior on the nonlinear part, \cite{Pensky:2007} propose Bayesian wavelet-smoothing of the log psd and recently, and \cite{Cadonna2017} modeled the log psd  as a mixture
of Gaussian distributions with frequency-dependent weights and mean functions.
%
\cite{Gangopadhyay:1999} modelled the log-spectral density function by  piecewise polynomials of low order between strategically placed knots on the support of the psd,  allowing the data to select the number and location of the knots.  A reversible jump Markov chain Monte Carlo (RJMCMC) \citep[RJMCMC;][]{Green:1995} algorithm was implemented to sample from the posterior distribution.

The P-spline prior on the spectral density function that we are going to propose is related to the 
Bernstein-Dirichlet process (BDP) prior  introduced by \cite{Choudhuri:2004} and combined with the Whittle likelihood.  The BDP prior  is a mixture of beta densities with weights induced by a Dirichlet process.  The number of components, which is a smoothing parameter, is given a discrete prior.  \cite{Edwards2018} extended the method by replacing the beta densities with B-spline densities. Similar to 
\cite{Gangopadhyay:1999}, the specification of the number and location of the B-spline densities is data-driven, however it avoids a RJMCMC algorithm by putting a second Dirichlet process prior  on the cdf that induces the inter-knot distances. 
Unlike the beta densities  which have global support on the interval [0,1], B-spline densities have local support, thereby increasing their flexibility and allowing better modelling of spectral densities with sharp peaks and abrupt changes.  As demonstrated in \citep{Edwards2018}, the B-spline--Dirichlet process (BspDP) prior outperforms the BDP prior in estimating complex psd's.  However, this flexibility comes at a high computational price.

%The Bernstein polynomial and the B-spline methods make use of Whittle's approximation to the Gaussian likelihood, known simply as Whittle likelihood \citep{Whittle:1957}.

With the aim of reducing the computational cost of the BspDP prior approach,  we suggest to use the idea of P-splines \cite{Eilers:1996} that combine a large but fixed number of  B-splines with a simple difference penalty that controls the degree smoothness of the spectral density. P-splines, i.e.\  equally-spaced B-splines,  have been successfully implemented for frequentist nonparametric regression but to the best of our knowledge not yet for Bayesian spectral density estimation. We suggest 
two novel Bayesian  approaches in this paper, both based on the Whittle likelihood and on a mixture of B-spline densities. Both approaches fix the  number of B-spline densities. 
The first approach uses equally spaced B-spline densities with a P-spline prior which we define in Section III. We demonstrate that this approach yields significant savings in computational time without sacrificing accuracy over the BspDP prior approach. The second approach has a particular  focus on spectral densities with sharp peaks. Instead of an equidistant spacing, the knots are spread based on the distribution of peaks in the periodogram. Thus the knot allocation is also data-driven as with the BspDP prior but the knots  remains fixed once
allocated. This novel knot allocation scheme avoids the computational complexity of the BspDP prior.  We show that with this knot allocation scheme, the estimates of spectral densities with spikes and abrupt changes improves significantly without an increase in computing time.

%Motivated by the flexibility of the B-spline prior and its associated high computational cost, in this work, we implement and test a particular case of the B-spline prior.  For this, we fix the %number of B-spline densities and place a P-spline prior on the spectral density instead.  This approach allows to penalize automatically the inclusion of B-spline densities and save significantly %computational time without sacrificing accuracy.  Actually, as will show, it outperforms the B-spline prior in our tests.  In addition, we propose a knot location scheme based on the %periodogram, which aims to deal with psd shapes that possess  abrupt peaks, without increasing computational time.  We show in an example how this scheme improves significantly the %estimates.

The paper is organized as follows: In order to set notation and allow for a direct comparison, we start by
reviewing the B-spline densities and the BspDP prior in Section II. Section III specifies the P-spline prior for the psd, the novel knot allocations scheme and the Gibbs sampler used to sample from the posterior distribution based on the P-spline prior and Whittle likelihood.  In Section IV, we test our P-spline-based approaches to spectral density estimation and compare results to those based on the BspDP prior in a simulation study and the application to two real time series. The paper concludes with a discussion of the relative merits of the new approaches.

\section{B-spline--Dirichlet process prior}

By an application of the Weierstrass theorem it can be seen that a mixture of beta densities with only integer parameters can uniformly approximate any continuous density  on the interval $[0,1]$ \citep{Choudhuri:2004}.  Let $G$ be a cumulative distribution function (cdf), with continuous density $g$, then 
\begin{align*}
\widehat{G}(\omega) &= \sum_{k=1}^{K} G \left( \dfrac{k-1}{K} , \dfrac{k}{K} \right] \beta(\omega; k, K-k+1)\\
&= \sum_{k=1}^{K} w_k \beta(\omega; k, K-k+1)
\end{align*}	  
converges uniformly to $G(\omega)$, where $G(u,v] = G(v) - G(u)$ and $\beta(\omega; a,b)$ is the beta density with parameters $a$ and $b$.
The Bernstein polynomial prior, which is used to describe a nonparametric prior for probability densities in the unit interval, is based on this approximation \citep{Petrone:1999a,Petrone:199b}.  \cite{Choudhuri:2004} proposed it as a prior on the spectral density function.

As shown by \cite{Perron:2001} for distributions on the interval [0,1], the mixture of B-spline distributions has better approximation properties than the mixture of beta distributions for a fixed number of components. In particular, the approximation error of the mixture of B-spline distributions 
can be made arbitrarily small by finding  suitable knot locations and increasing the number of knots.
%the mixture of beta distributions is unable to provide a full coverage of the space of cdfs on $[0,1]$.  They defined a loss function and showed that this cannot be arbitrarily small under this %approach.  Alternatively, they showed that if the beta distributions are replaced by B-spline distributions of fixed order (shown for order~2) with variable knots, the loss function can be %arbitrarily small by increasing the number of knots.  The adequacy of the approximation is a function of the number of knots.  
This work motivated \cite{Edwards2018} to propose the BspDP prior on the spectral density function.

@@

The \textit{B-spline densities}, i.e.\ normalized B-spline functions, are defined by
\begin{align*}
b_{k,r}(\omega; \bm{\xi}) = \left(r+1\right) \dfrac{B_{k,r}(\omega;\bm{\xi})}{\xi_{k+r} - \xi_{k-1}}.
\end{align*}

@@

\subsection*{Prior description}
The BspDP prior is a mixture of B-spline densities defined as
\begin{align}
s_r(\omega; K,\bm{w}_K, \bm{\xi}) = \sum_{k=1}^{K} w_{k,K} b_{k,r}(\omega;\bm{\xi}),
\label{eq:bs_prior}	
\end{align}
where $K$ is the number of B-spline densities  $b_{k,r}(\cdot)$ of fixed degree $r$, $\bm{w}_K=(w_{1,K},\cdots, w_{K,K})$ the weight vector, and $\bm{\xi}$ is the knot sequence.

@@

Assuming that the weights are induced by a cdf $G$ on $[0,1]$, \cite{Choudhuri:2004} proposed a Dirichlet process as a prior for $G$.  This approach avoids the potential difficulties caused by  jumps between the parameter spaces of different dimensions in  RJMCMC.  \cite{Edwards2018} followed this approach, but in addition they assumed that the $K-r$ internal knot differences are induced by a cdf $H$ on $[0,1]$ with further Dirichlet process prior, that is
\begin{align*}
\Delta_{j} = \xi_{j+r} - \xi_{j+r-1} = H\left(\dfrac{j-1}{K-r}, \dfrac{j}{K-r} \right],
\end{align*}
for $j=1,\dots,K-r$.

Therefore, the BspDP prior on the spectral density function parametrized in terms of $K$, $G$, and $H$ is given by
\begin{align*}
s_r(\omega;K,G,H) = \sum_{k=1}^{K} G \left( \dfrac{k-1}{K} , \dfrac{k}{K} \right] b_{k,r}(\omega;H).
\end{align*}

The BspDP prior is a mixture of B-spline densities with local support, unlike the BDP prior which uses beta densities with  support on the whole unit interval.  This feature allows the B-spline method more flexibility for spectral density estimation \citep{Edwards2018}.

When there are no internal knots, the B-spline basis coincides with the Bernstein polynomial basis.  Therefore, the BspDP  prior can be considered  a generalization of the BDP prior.

@@

The BspDP prior discussed above has been defined for the unit interval $[0,1]$.  However, the spectral density function $f(\cdot)$ is defined on the interval $[0,\pi]$, therefore its prior is reparametrized and defined as
\begin{align*}
f(\pi \omega) = \tau \times s_r(\omega;K,G,H), \quad \omega \in [0,1],
\end{align*}
where $\tau = \int_{0}^{1}f(\tau \omega)\text{d}\omega$ is the normalizing constant.

To sum up, the BspDP prior has the following components:
\begin{itemize}
	\item $G \sim \text{DP}(M_G, G_0)$, where $M_G>0$ is the concentration parameter and $G_0$ is the base distribution.  It produces the weights for the $K$ B-spline densities.  
	\item $H \sim \text{DP}(M_H, H_0)$, where $M_H>0$ is the concentration parameter and $H_0$ is the base distribution.  It produces the knot locations determining the shape and location of the B-spline densities.
	\item $p(K) \propto \text{exp}(-\theta_K K^2)$ for $K=1,\dots,K_{\text{max}}$.  $K$ is the number of B-spline densities used in Eq. \eqref{eq:bs_prior}.  It plays the role of smoothing parameter, that is smaller values implies smoother spectral density estimates.  $K_{\text{max}}$ must be limited  for computational reasons.  In practice, pilot runs are used to determine it.
	\item $\tau \sim \text{Inverse-Gamma}(\alpha_{\tau},\beta_{\tau})$ is the normalizing constant.  $\alpha_{\tau}$ and $\beta_{\tau}$ are the shape and scale parameters, respectively.
\end{itemize}

%\subsection*{Implementation discussion}

The BspDP prior proposed by \cite{Edwards2018} involves two  Dirichlet processes, one  placed on $G$ inducing  the weights of the B-spline densities and the other on  $H$ for the knot locations.  In practice, these are implemented using   the  stick-breaking representation \citep{Sethuraman:1994}, for details see \cite{Edwards2018}. 
% This method consists basically in a representation of the Dirichlet process as a sequential division of the unit interval infinite times according to a beta distribution. 
For computational reasons, the number of components of this infinite series representation of $G$ and $H$ is limited to a large but finite positive integer, that is $L_G$ and $L_H$, respectively.  These parameters control the quality of the approximations.  Larger values improve the approximation, but increase the number of calculations and consequently the computation time.

As shown by \cite{Edwards2018}, the approach based on the  BspDP prior outperforms the BDP prior in estimating spiky spectral densities.  In the case of smooth
spectral densities, there is no a significant difference in the performance of these two approaches.  However, in practice, the true psd function is unknown, which makes it necessary to use   methods such as the BspDP that will work well in all conceivable cases.

To implement the BDP prior, one only has to  calculate the beta densities once and they can be stored and re-utilized across MCMC steps.  On the other hand, the BspDP prior requires the calculation of the B-spline densities at each iteration, since these vary due to varying knot locations.  Moreover,  $2\!\times\!(L_G + L_H)$ calculations are needed in the truncated stick-breaking representation of the two Dirichlet processes  at each iteration %(see Eq. \eqref{eq:prior})
resulting in a significantly increased complexity of  the  BspDP algorithm.

This has prompted us to explore the performance of two new algorithms, both based on the P-spline prior algorithm. Both use a fixed number of B-spline densities and knot locations, avoiding the Dirichlet processes and the recalculation of the B-spline densities. Whereas the first approach uses  equidistant knots, the second algorithm  proposes a new fast knot allocation strategy. Both methods preserve the flexibility of the B-spline densities and their potential ability to estimate spectral densities with sharp peaks and abrupt changes.

@@

Fixing the number of B-spline densities $K$ in Eq. \eqref{eq:bs_prior}, the prior for the unit interval $[0,1]$ is defined as
\begin{align*}
s_r(\omega;\bm{w}, \bm{\xi}) = \sum_{k=1}^{K} w_{k} b_{k,r}(\omega;\bm{\xi}),	
\end{align*}
where $\bm{w} = (w_1,\dots,w_K)$.

$\bm{\xi}$ contains fixed equidistant internal knots on $[0,1]$ in our basic P-spline prior. A different  allotment of knots is described in  section ``knot allocation strategy" .
Therefore, the P-spline prior on the spectral density function is given by
\begin{align*}
f(\pi \omega) = \tau \times s_r(\omega;\bm{w}, \bm{\xi}), \quad \omega \in [0,1],
\end{align*}
where $\tau = \int_{0}^{1}f(\tau \omega)\text{d}\omega$ is the normalizing constant.
